---
title: "Reflected Ornstein–Uhlenbeck (ROU) Process"
author: "Brolin O'Connell"
date: "2025-06-17"
output:
  pdf_document:
    toc: true
    toc_depth: '2'
  html_document:
    toc: true
    toc_depth: 2
    theme: flatly
  bookdown::pdf_document2:
    latex_engine: xelatex
    number_sections: true
    toc: true
    toc_depth: 2
    fig_caption: true
    fig_width: 6.5
    fig_height: 4.2
    fig_pos: H
    keep_tex: true
header-includes:
- \usepackage[font=small,labelfont=bf]{caption}
- \usepackage{float}
- \usepackage{titlesec}
- \titlespacing*{\section}{0pt}{12pt plus 4pt minus 2pt}{6pt plus 2pt minus 2pt}
- \titlespacing*{\subsection}{0pt}{10pt plus 4pt minus 2pt}{4pt plus 2pt minus 2pt}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 6.5,
  fig.height = 3.85,
  fig.align = "center",
  out.width = "100%"
)

library(tidyverse)
library(Sim.DiffProc)
```

# Introduction

The Ornstein–Uhlenbeck (OU) process is a classical mean-reverting stochastic process used in a variety of applications including physics, finance, and queueing theory. In this document, we simulate and analyze both the standard OU process and its **reflected** form (ROU) in one and two dimensions, examining key properties such as stationarity, ergodicity, and convergence.

```{=latex} \vspace{10pt} ```

---

# 1. Ornstein–Uhlenbeck Process (1D)

The 1D OU process is governed by the stochastic differential equation:

$$
dX_t = -\theta (X_t - \mu) \, dt + \sigma \, dW_t \tag{1}
$$

Where:
- $X_t$: state at time $t$
- $\mu$: long-run mean
- $\theta > 0$: mean reversion speed
- $\sigma$: volatility
- $dW_t$: standard Brownian motion increment

This process is stationary and has a well-defined limiting distribution:

$$
\text{Var}(X_t) = \frac{\sigma^2}{2\theta} \tag{2}
$$

(see Lemma 7, Huang & Pang, 2022)

```{r ou-paths, echo=FALSE}
# Parameters
theta <- 0.7
mu <- 1.0
sigma <- 0.3
x0 <- 0.0
T <- 10
N <- 1000
dt <- T / N
time <- seq(0, T, length.out = N)
n_paths <- 5

# OU path simulation
simulate_ou <- function() {
  x <- numeric(N)
  x[1] <- x0
  for (i in 2:N) {
    dw <- rnorm(1, mean = 0, sd = sqrt(dt))
    x[i] <- x[i-1] + theta * (mu - x[i-1]) * dt + sigma * dw
  }
  return(x)
}

ou_paths <- replicate(n_paths, simulate_ou())
ou_df <- as.data.frame(ou_paths)
ou_df$Time <- time
ou_long <- pivot_longer(ou_df, -Time, names_to = "Path", values_to = "X")

# Plot
ggplot(ou_long, aes(x = Time, y = X, color = Path)) +
  geom_line(size = 1) +
  theme_minimal(base_size = 14) +
  labs(title = "Ornstein–Uhlenbeck Process (1D)",
       x = "Time",
       y = expression(X[t])) +
  theme(legend.position = "bottom")
```
```{=latex} \vspace{10pt} ```

---

# 2. Stationarity and Ergodicity

The OU process is **second-order stationary**: its mean and variance are time-invariant, and autocovariance depends only on lag. It is also **ergodic**, meaning:

$$
\lim_{T \to \infty} \frac{1}{T} \int_0^T X_t \, dt = \mathbb{E}[X_t] = \mu \tag{3}
$$

(see Theorem 2.4, Huang & Pang, 2022)
```{=latex} \vspace{10pt} ```
---

# 3. Multivariate OU Process (2D)

We now extend the OU process to two dimensions, where the state vector \( \mathbf{X}_t = (X_t, Y_t) \in \mathbb{R}^2 \) evolves as:

$$
d\mathbf{X}_t = -\Theta(\mathbf{X}_t - \boldsymbol{\mu})\, dt + \Sigma\, d\mathbf{W}_t \tag{4}
$$

Where:
- $\Theta$: mean reversion matrix
- $\Sigma$: diffusion matrix
- $\boldsymbol{\mu}$: long-run mean vector
- $d\mathbf{W}_t$: vector of independent Brownian motions

```{r ou-2d, echo=FALSE}
set.seed(123)
Theta <- matrix(c(1.0, 0.3, 0.2, 0.8), nrow = 2, byrow = TRUE)
mu_vec <- c(1, 1.5)
Sigma <- matrix(c(0.2, 0.05, 0.05, 0.3), nrow = 2, byrow = TRUE)
x_init <- c(0, 0)
T <- 10
dt <- 0.01
N <- T / dt
time <- seq(0, T, length.out = N)

X <- matrix(0, nrow = N, ncol = 2)
X[1, ] <- x_init
for (i in 2:N) {
  dW <- rnorm(2, mean = 0, sd = sqrt(dt))
  drift <- -Theta %*% (X[i-1, ] - mu_vec)
  diffusion <- Sigma %*% dW
  X[i, ] <- X[i-1, ] + drift * dt + diffusion
}

df_2d <- data.frame(Time = time, X_t = X[,1], Y_t = X[,2])
df_long <- pivot_longer(df_2d, cols = c("X_t", "Y_t"), names_to = "Component", values_to = "Value")

ggplot(df_long, aes(x = Time, y = Value, color = Component)) +
  geom_line(size = 1) +
  theme_minimal(base_size = 14) +
  labs(title = "2D Ornstein–Uhlenbeck Process Components",
       x = "Time", y = "Value") +
  theme(legend.title = element_blank())
```
```{=latex} \vspace{10pt} ```
---

### 4. Reflected Ornstein–Uhlenbeck Process (1D)

We now simulate a **reflected OU process** in 1D, where the process is constrained to remain in \( \mathbb{R}_+ \). This imposes a Skorokhod reflection at the boundary \( X_t = 0 \), approximated here by taking the absolute value of a standard OU trajectory.
```{r, echo=FALSE}
library(Sim.DiffProc)
library(tidyverse)

# Parameters
theta <- 0.7
mu <- 0.0
sigma <- 0.3
T <- 10
N <- 1000
x0 <- -1.5  # Start below 0 for visible reflection

# Simulate OU process
y_path <- as.numeric(snssde1d(
  drift = expression(theta * (mu - x)),
  diffusion = expression(sigma),
  M = 1, N = N, T = T, x0 = x0
)$X)

time <- seq(0, T, length.out = N + 1)
x_reflected <- abs(y_path)

# Dataframe for plotting
df <- tibble(
  Time = rep(time, 2),
  Value = c(y_path, x_reflected),
  Type = rep(c("Original OU", "Reflected |OU|"), each = N + 1)
)

# Plot
ggplot(df, aes(x = Time, y = Value, color = Type, linetype = Type)) +
  geom_line(linewidth = 1.2) +
  geom_hline(yintercept = 0, color = "black", linewidth = 0.5) +
  theme_minimal(base_size = 14) +
  scale_color_manual(values = c("Original OU" = "darkred", "Reflected |OU|" = "steelblue")) +
  labs(title = "True Reflection of Ornstein–Uhlenbeck Process at x = 0",
       x = "Time", y = expression(X[t]), color = "Process", linetype = "Process")
```

```{=latex} \vspace{10pt} ```


---


### 5. Local Time Approximation

Following Huang & Pang (2022), the reflected process introduces a local time \( L_t \), which increases only when \( X_t = 0 \). To approximate this numerically, we accumulate the negative excursions of the unreflected path:

```{r, echo=FALSE}
# Approximate local time as accumulated negative excursions
dt <- T / N
L_t <- cumsum(ifelse(y_path < 0, abs(y_path), 0)) * dt

# Add to the plot
df_local <- tibble(Time = time, LocalTime = L_t)

ggplot(df_local, aes(x = Time, y = LocalTime)) +
  geom_line(color = "purple", linewidth = 1.2) +
  theme_minimal(base_size = 14) +
  labs(title = "Accumulated Local Time Process L(t)",
       x = "Time", y = expression(L[t]))
```

```{=latex} \vspace{10pt} ```

---


### 6. Reflected OU Process in \( \mathbb{R}_+^2 \)

We extend the OU system to 2D and impose reflection into the nonnegative orthant by projecting any negative value to zero.

```{r, echo=FALSE}
library(tidyverse)

# Parameters
set.seed(123)
T <- 10
dt <- 0.01
N <- T / dt
time <- seq(0, T, length.out = N)

# Initial state
x_init <- c(-1, -2)  # Start below 0 for visibility

# Mean vector
mu_vec <- c(1, 1.5)

# Mean reversion matrix (Theta)
Theta <- matrix(c(1.0, 0.3,
                  0.2, 0.8), nrow = 2, byrow = TRUE)

# Volatility matrix (Sigma)
Sigma <- matrix(c(0.2, 0.05,
                  0.05, 0.3), nrow = 2, byrow = TRUE)

# Simulate standard 2D OU process
X <- matrix(0, nrow = N, ncol = 2)
X[1, ] <- x_init

for (i in 2:N) {
  dW <- rnorm(2, mean = 0, sd = sqrt(dt))
  drift <- -Theta %*% (X[i-1, ] - mu_vec)
  diffusion <- Sigma %*% dW
  X[i, ] <- X[i-1, ] + drift * dt + diffusion
}

# Reflect: componentwise absolute value
X_reflected <- abs(X)

# Combine both into long format
df_original <- data.frame(Time = time, X_t = X[,1], Y_t = X[,2], Type = "Original")
df_reflected <- data.frame(Time = time, X_t = X_reflected[,1], Y_t = X_reflected[,2], Type = "Reflected")

df_combined <- bind_rows(df_original, df_reflected) %>%
  pivot_longer(cols = c("X_t", "Y_t"), names_to = "Component", values_to = "Value")

# Plot
ggplot(df_combined, aes(x = Time, y = Value, color = interaction(Type, Component), linetype = Type)) +
  geom_line(linewidth = 1) +
  geom_hline(yintercept = 0, color = "black", linewidth = 0.4) +
  theme_minimal(base_size = 14) +
  labs(
    title = "Original vs Reflected 2D OU Process (|Xt| and |Yt|)",
    x = "Time", y = "Value",
    color = "Process + Component",
    linetype = "Path Type"
  )
```

```{=latex} \vspace{10pt} ```

--- 

### 7. Autocorrelation Function (ACF)

```{r, echo=FALSE}
# Long OU path for ACF
set.seed(42)
acf_path <- simulate_ou()
acf(acf_path, lag.max = 50, main = "Autocorrelation of OU Process")
```

OU processes exhibit exponentially decaying autocorrelations: \( \rho(\tau) = e^{-\theta \tau} \).

```{=latex} \vspace{10pt} ```

--- 

### 8. Stationary Distribution Convergence

We compare empirical distribution at time \( T \) to the theoretical stationary normal:

```{r, echo=FALSE}
# Stationary distribution simulation
n_samples <- 10000
final_values <- replicate(n_samples, simulate_ou()[N])
df_final <- tibble(Value = final_values)

# Theoretical parameters
mu_theoretical <- mu
sd_theoretical <- sqrt(sigma^2 / (2 * theta))

# Plot
 ggplot(df_final, aes(x = Value)) +
  geom_histogram(aes(y = ..density..), bins = 60, fill = "#1f78b4", alpha = 0.6) +
  stat_function(fun = dnorm, args = list(mean = mu_theoretical, sd = sd_theoretical),
                color = "darkred", linewidth = 1.2) +
  theme_minimal(base_size = 15) +
  labs(title = "Empirical Stationary Distribution vs Theoretical Normal",
       x = "Final Value of Xₜ", y = "Density") +
  annotate("text", x = mu + 0.5, y = 1,
           label = paste0("N(", mu, ", ", round(sd_theoretical^2, 3), ")"),
           color = "darkred", size = 4)
```

> Huang & Pang (2022) confirm the convergence of the ROU process toward a stationary distribution under bounded local time and ergodicity (Theorem 2.4).








